### 2019-04-27 Thesis/pmonads.tex
Monthly commit.

\documentclass{article}

\input{../common/preamble}
\begin{document}
\section{Parametric monads}
\label{ChapParametricMonads}

@@

\subsection{The \Mellies category}

@@

\subsection{The category $\C/\X$}

@@

\subsection{Lax $2$-colimits}

In the previous section, we approached the category $\C/\X$ via the \Mellies category.  
In this section, we will show that $\C/\X$ has important properties in its own right: namely that it is a certain lax $2$-colimit in $\Cat$.  
We will briefly return to the \Mellies category in Section \ref{SecPromonads}, when we will show why it is a natural idea to consider an $[\X,\Set]$-enriched category.

\begin{definition}[\cite{StreetTwoConstructions}]
  Let $\C,\D$ be bicategories.  
  Then a \emph{lax functor} $F\from\C \to \D$ is given by
  \begin{itemize}
    \item a map $F$ from the objects of $\C$ to the objects of $\D$,
    \item for each pair $a,b$ of objects of $\C$, a functor $F$ from $\C(a,b)$ to $\D(F(a),F(b))$,
    \item for each triple $a,b,c$ of objects of $\C$, a transformation $m_{f,g}\from F(f);F(g) \to F(f;g)$ natural in $f\from a \to b$, $g\from b \to c$ and
    \item for each object $a$ of $\C$, a $2$-cell $e_a\from \id_{Fa} \Rightarrow F(\id_a)\from Fa \to Fa$
  \end{itemize}
  such that for all tuples $a,b,c,d$ of objects and all morphisms $f\from a \to b$, $g\from b \to c$, $h \from c \to d$, the following diagrams commute.
  \begin{mathpar}
    \begin{tikzcd}[column sep=60pt]
      (F(f);F(g));F(h) \arrow[r, "{\assoc_{F(f),F(g),F(h)}}"] \arrow[d, "{m_{f,g};F(h)}"']
        & F(f);(F(g);F(h)) \arrow[d, "{F(f);m_{g,h}}"] \\
      F(f;g);F(h) \arrow[d, "{m_{f;g,h}}"']
        & F(f);F(g;h) \arrow[d, "{m_{f,g;h}}"] \\
      F((f;g);h) \arrow[r, "{F(\assoc_{f,g,h})}"]
        & F(f;(g;h))
    \end{tikzcd}
    \and
    \begin{tikzcd}
      Ff \arrow[r, "\lunit_{Ff}"] \arrow[d, "F\lunit_f"']
        & \id_{Fa} ; Ff \arrow[d, "e_a;Ff"] \\
      F(\id_a;f)
        & F\id_a;Ff \arrow[l, "{m_{\id_a,f}}"']
    \end{tikzcd}
    \and
    \begin{tikzcd}
      Ff \arrow[r, "\runit_{Ff}"] \arrow[d, "F\runit_f"']
        & Ff ; \id_{Fb} \arrow[d, "Ff;e_b"] \\
      F(f;\id_b)
        & Ff;F\id_b \arrow[l, "{m_{f,\id_b}}"']
    \end{tikzcd}
  \end{mathpar}
\end{definition}
\begin{example}
  If $\C = \BB \X$, $\D = \BB \YY$ are the delooping bicategories of monoidal categories $\X, \YY$ -- i.e., $\X$ and $\YY$ considered as bicategories with a single object -- then a lax functor $\C \to \D$ is the same thing as a lax monoidal functor $\X \to \YY$.

@@

\subsection{Lax natural transformations and functoriality of lax colimits}

@@

\subsection{Lax $2$-colimits in $\Cat$}

@@

\subsection{Examples of lax $2$-colimits in $\Cat$}

@@

This is significant, because it tells us that the category $\C/\X$ always satisfies a `factorization result' akin to those in \cite{SamsonGuyIAActive} and \cite{mcCHFiniteND}: if $f \from a \to b$ is a morphism in $\C/\X$, given by some morphism $\tilde{f}\from a \to x.b$ in $\C$, then we may write $f$ as the composite
\[
  a \xrightarrow{J(\tilde{f})} x.b \xrightarrow{\phi_{x,b}} b
  \]
in $\C$.  
This means that if we have a definability result for the category $\C$ -- for example, that every compact morphism in $\C$ is definable in some language $\L$ -- then we can automatically get a definability result for the category $\C/\X$ -- for example, that every compact morphism in $\C/\X$ (i.e., every morphism $a\to b$ in $\C/\X$ that is given by a compact morphism $a \to x.b$ in $\C$) is definable in the language $\L + \Phi_{x,a}$, where $\Phi_{x,a}$ is a new family of primitives in the language whose denotations are given by the $\phi_{x,a}$.
This is the same purpose that we used Proposition \ref{pKleisli} for when proving Theorem \ref{TheKleisliFullAbstraction}.

\subsection{Finite products distribute over lax colimits in $\Cat$}

@@

\subsection{Monoidal structure of $\C/\X$}

@@

\subsection{Symmetric monoidal structure of $\C/\X$}

@@

\subsection{Symmetric monoidal closed structure of $\C/\X$}

@@

\subsection{Cartesianness of the monoidal structure}

Suppose that a symmetric monoidal category $\X$ acts via a symmetric monoidal action on a Cartesian category $\C$.  
In this section we will consider what properties of the action we need in order to ensure that the induced monoidal structure on $\C/\X$ is Cartesian.

Since $\C$ is Cartesian, the diagonal map in $\C$ gives every object a natural comonoid structure:
\[
  a \xrightarrow{\Delta_a} a \times a\,.
  \]
Since the tensor product of objects in $\C/\X$ is given by the same object as in $\C$, we can lift this comonoid structure through the functor $J$ to give us a natural comonoid structure on each object of $\C/\X$:
\[
  a \xrightarrow{J\Delta_a} a \tensor a\,.
  \]
We use the symbol $\tensor$ rather than $\times$ in $\C/\X$, since this is not necessarily a category-theoretic product.

If every morphism $f\from a \to b$ in $\C/\X$ is a comonoid homomorphism with respect to this comonoid structure; i.e., if it makes the diagram
\[
  \begin{tikzcd}
    a \arrow[r, "f"] \arrow[d, "J\Delta_a"']
      & b \arrow[d, "J\Delta_b"] \\
    a \tensor a \arrow[r, "f\tensor f"]
      & b \tensor b
  \end{tikzcd}
  \]
commute, then we may identify $\C/\X$ with a full subcategory of its own category of comonoids, closed under tensor product.  
Since the category of comonoids in a monoidal category is automatically Cartesian, this will tell us that $\C/\X$ is Cartesian.

What does it mean for every morphism in $\C/\X$ to be a comonoid homomorphism?  
Firstly, since $\C$ itself is Cartesian, every morphism in $\C$ is a comonoid homomorphism with respect to the diagonal, and it follows that every morphism of the form $Jf \from a \to b$ (where $f\from a \to b$ is a morphism in $\C$) is a comonoid homomorphism in $\C/\X$.  
Since every morphism in $\C/\X$ may be written as the composite of a morphism of the form $Jf$ with a morphism of the form $\phi_{x,b}$, it will suffice to show that the morphisms of the form $\phi_{x,a}$ are comonoid homomorphisms in $\C/\X$; i.e., that they make the following diagram commute.
\[
  \begin{tikzcd}[column sep=40pt]
    x.a \arrow[r, "{\phi_{x,a}}"] \arrow[d, "J\Delta_{x.a}"']
      & a \arrow[d, "J\Delta_a"] \\
    x.a \tensor x.a \arrow[r, "{\phi_{x,a}\tensor\phi_{x,a}}"]
      & a \tensor a
  \end{tikzcd}
  \]
Let us compute the two arms of this square directly in $\C$.  
The composite along the top right is given in $\C$ by the morphism
\[
  x.a \xrightarrow{x.\Delta_a}
  x.(a\times a)\,,
  \]
while that along the bottom right is given by the composite
\[
  x.a \xrightarrow{\Delta_{x.a}}
  x.a \times x.a \xrightarrow{m_{x,x,a,a}}
  (x\tensor x) . (a \times a)\,.
  \]
Then a sufficient condition for these two morphisms to be equal in $\C/\X$, and therefore for the category $\C/\X$ to be Cartesian, is that for every object $x$ of $\X$ there should be a morphism $f \from x \to x\tensor x$ that makes the following diagram commute.
\[
  \begin{tikzcd}
    x.a \arrow[dr, "f.\Delta_a"] \arrow[d, "\Delta_{x.a}"']
      & \\
    x.a \times x.a \arrow[r, "{m_{x,x,a,a}}"]
      & (x\tensor x). (a\times a)
  \end{tikzcd}
  \]
If $\X$ itself is Cartesian, then we have a natural choice of morphism $x \to x \times x$, but choosing $f$ to be the diagonal map does not automatically make the diagram above commute.  
For example, the Cartesian product is not the category-theoretic product in the category of sets and relations, even though this category is the Kleisli category for the symmetric monoidal monad $\powerset \from \Set \to \Set$, and may be identified with the category $\Set/1$, where $1$ and $\Set$ are both Cartesian.  
In this case, the diagram
\[
  \begin{tikzcd}
    \powerset(A) \arrow[dr, "\powerset(\Delta_A)"] \arrow[d, "\Delta_{\powerset(A)}"']
      & \\
    \powerset(A) \times \powerset(A) \arrow[r, "{m_{A,A}}" xshift=-3pt]
      & \powerset(A \times A)
  \end{tikzcd}
  \]
does not commute, the arrow at the top right sending a set $X\subseteq A$ to the set $\{(x,x) \suchthat x \in X\}$ while the composite at the bottom left sends $X$ to the set $\{(x,y)\suchthat x,y\in X\}$.

An alternative sufficient condition for $\C/\X$ to be Cartesian is for the morphism mediating between the two arms of the square to go in the other direction.  
In this case, we require that for each object $x$ of $\X$ there should be a morphism $g \from x\tensor x \to x$ that makes the following diagram commute.
\[
  \begin{tikzcd}
    x.a \arrow[r, "x.\Delta_a"] \arrow[d, "\Delta_{x.a}"']
      & x.(a\times a) \\
    x.a \times x.a \arrow[r, "{m_{x,x,a,a}}"]
      & (x \tensor x).(a \times a) \arrow[u, "g.(a\times a)"']
  \end{tikzcd}
  \]
This will be particularly useful for reader actions.  
Indeed, suppose that $j\from \X \to \C$ is an oplax symmetric monoidal functor between symmetric monoidal categories, where $\C$ is Cartesian closed.  
The functor $j$ induces a reader action of $\oppcat\X$ on $\C$ (i.e., $x.a = jx \to a$), and we want the category $\C/\oppcat\X$ to be Cartesian.  
For this, it will be sufficient to find, for each object $x$ of $\X$, a morphism $f \from x \to x\tensor x$ such that the following diagram commutes.
\[
  \begin{tikzcd}
    jx \to a \arrow[rr, "jx \to \Delta_a"] \arrow[d, "\Delta_{jx \to a}"']
      &[-10pt]
        &[+10pt] jx \to (a \times a) \\
    (jx \to a) \times (jx \to a) \arrow[r]
      & (jx \times jx) \to (a \times a) \arrow[r, "m^j_x \to (a\times a)"]
        & j(x\tensor x) \to (a \times a) \arrow[u, "jf \to (a \times a)"']
  \end{tikzcd}
  \]
For this to commute, it is sufficient that $jf$ should be equal to the following composite.
\[
  jx \xrightarrow{\Delta_{jx}}
  jx \times jx \xrightarrow{m^j_{x,x}}
  j(x \tensor x)\,.
  \]
We have proved:

@@

  Suppose that for every object $x$ of $\X$, there is a morphism $f \from x \to x\tensor x$ in $\X$ such that $jf=\Delta_{jx};m^j_{x,x}$.  
  Then $\C/\oppcat \X$ is Cartesian closed.
  \label{TheCartesianClosedCx}
\end{theorem}

In the next section, we will investigate a class of reader actions, and give a more concrete property for the category $\C/\oppcat\X$ to be Cartesian closed.

\section{Reader actions on $\Set$}

\subsection{Colimits of actions are monoidal functors}

@@

        \colim_{t,z\from \X}\limits t.(a \tensor b) \tensor z.c \arrow[d, "{\Colim m_{t,z,a\tensor b,c}}"']

@@

              \& \colim_{x,t\from \X}\limits x.a \tensor t.(b\tensor c) \arrow[d, "{\Colim m_{x,t,a,b\tensor c}}"] \\
        \colim_{t,z\from \X}\limits (t\tensor z).((a \tensor b)\tensor c) \arrow[d, hookrightarrow]
          \& \colim_{x,y,z\from\X}\limits ((x\tensor y) \tensor z).((a \tensor b) \tensor c) \arrow[r, "\Colim\assoc.\assoc" yshift=5pt] \arrow[l, hookrightarrow]
            \& \colim_{x,y,z\from\X}\limits (x \tensor (y\tensor z)).(a \tensor (b \tensor c)) \arrow[r, hookrightarrow]
              \& \colim_{x,t\from \X}\limits (x \tensor t).(a \tensor (b \tensor c)) \arrow[d, hookrightarrow] \\
        \colim_{u\from \X}\limits u.((a\tensor b) \tensor c) \arrow[rrr, "\Colim u.\assoc"]

@@

\subsection{Reader actions on $\Set$ vs change of base}

@@

\begin{proposition}
  Let $j \from \X \to \Set$ be an oplax symmetric monoidal functor between symmetric monoidal categories.
  Let $\oppcat\X$ act on $\Set$ via the reader action induced by $j$.
  Then the \Mellies category $\Mell_{\oppcat\X}\Set$ is isomorphic to the category obtained via base change along the action regarded as a functor
  \[
    \blank.\blank\from \Set \to [\X,\Set]\,.
    \]
  \label{PropMelliesVsBaseChange}
\end{proposition}
\begin{proof}
  The objects of both categories are sets.  
  The morphism objects in the \Mellies category are given by
  \[
    \Mell_\X\Set(A,B)(x) = [A,[jx,B]]\,,
    \]
  while those in the base-changed category are given by
  \[
    (\blank.\blank)_*\Set(A,B)(x) = [jx, [A,B]]\,,
    \]
  and these may be related by the symmetry isomorphism
  \[
    s_{jx,A,B} \from [A,[jx,B]] \to [jx, [A,B]]\,.
    \]
  We need to show that this preserves composition of morphisms.  
  Recall that composition in the \Mellies category is given by
  \begin{IEEEeqnarray*}{Cl}
    & \int^{y,z} [A,[jy,B]] \times [B,[jz,C]] \times \X(y\tensor z,x) \\
    \to &
    \int^{y,z} [A,[j(y\tensor z),C]]\times \X(y\tensor z,x) \\
    \cong &
    [A,[jx,C]]\,,
  \end{IEEEeqnarray*}
  where the first arrow is induced via the \Mellies composition, while composition in the base changed category is given by
  \begin{IEEEeqnarray*}{Cl}
    &\int^{y,z}[jy,[A,B]] \times [jz,[B,C]] \times \X(y\tensor z,x) \\
    \xrightarrow{\makebox[146pt]{$\int^{y,z} m_{jy,jz,[A,B],[B,C]}\times \X(y\tensor z,x)$}} &
    \int^{y,z}[jy \times jz,[A,B]\times [B,C]] \times \X(y,\tensor z,x) \\
    \xrightarrow{\makebox[146pt]{$\int^{y,z} [m^j_{y,z}, ;] \times \X(y\tensor z,x)$}} &
    \int^{y,z}[j(y \tensor z),[A,C]]\times \X(y\tensor z,x) \\
    \cong & 
    [jx,[A,C]]\,.
  \end{IEEEeqnarray*}
  where $;$ is the internal composition in $\Set$.

@@

\subsection{From monoidal endofunctors on $\Set$ to reader actions}

@@

Now recall that in order to apply the second part of Theorem \ref{TheCartesianClosedCx}, and deduce that $\Set/\oppcat\X$ is Cartesian closed, we must prove that for every object $x$ of $\X$, there is a morphism $f \from x \to x \tensor x$ such that $jf = \Delta_{jx};m^j_x$.  
This is particularly useful in our case, since the functor $\X \to \Set$ is faithful.

>>>>

\documentclass{report}[11pt]

\input{../common/preamble}
\usepackage{lua-visual-debug}
\begin{document}
\chapter{Parametric monads}
\label{ChapParametricMonads}

@@

\section{The \Mellies category}

@@

\section{The category $\C/\X$}

@@

\section{Lax $2$-colimits}

In the previous chapter, we approached the category $\C/\X$ via the \Mellies category.  
In this chapter, we will show that $\C/\X$ has important properties in its own right: namely that it is a certain lax $2$-colimit in $\Cat$.  
We will briefly return to the \Mellies category in chapter \ref{SecPromonads}, when we will show why it is a natural idea to consider an $[\X,\Set]$-enriched category.

\begin{definition}[\cite{StreetTwoConstructions}]
  Let $\C,\D$ be bicategories.  
  A \emph{lax functor} $F\from\C \to \D$ is given by
  \begin{itemize}
    \item a map $F$ from the objects of $\C$ to the objects of $\D$,
    \item for each pair $a,b$ of objects of $\C$, a functor $F$ from $\C(a,b)$ to $\D(F(a),F(b))$,
    \item for each triple $a,b,c$ of objects of $\C$, a transformation $m_{f,g}\from F(f);F(g) \to F(f;g)$ natural in $f\from a \to b$, $g\from b \to c$ and
    \item for each object $a$ of $\C$, a $2$-cell $e_a\from \id_{Fa} \Rightarrow F(\id_a)\from Fa \to Fa$
  \end{itemize}
  such that for all tuples $a,b,c,d$ of objects and all morphisms $f\from a \to b$, $g\from b \to c$, $h \from c \to d$, the following diagrams commute.
  \begin{mathpar}
    \begin{tikzcd}[column sep=60pt]
      (F(f);F(g));F(h) \arrow[r, "{\assoc_{F(f),F(g),F(h)}}"] \arrow[d, "{m_{f,g};F(h)}"']
        & F(f);(F(g);F(h)) \arrow[d, "{F(f);m_{g,h}}"] \\
      F(f;g);F(h) \arrow[d, "{m_{f;g,h}}"']
        & F(f);F(g;h) \arrow[d, "{m_{f,g;h}}"] \\
      F((f;g);h) \arrow[r, "{F(\assoc_{f,g,h})}"]
        & F(f;(g;h))
    \end{tikzcd}
    \and
    \begin{tikzcd}
      Ff \arrow[r, "\lunit_{Ff}"] \arrow[d, "F\lunit_f"']
        & \id_{Fa} ; Ff \arrow[d, "e_a;Ff"] \\
      F(\id_a;f)
        & F\id_a;Ff \arrow[l, "{m_{\id_a,f}}"']
    \end{tikzcd}
    \and
    \begin{tikzcd}
      Ff \arrow[r, "\runit_{Ff}"] \arrow[d, "F\runit_f"']
        & Ff ; \id_{Fb} \arrow[d, "Ff;e_b"] \\
      F(f;\id_b)
        & Ff;F\id_b \arrow[l, "{m_{f,\id_b}}"']
    \end{tikzcd}
  \end{mathpar}
\end{definition}
\begin{example}
  If $\C = \BB \X$, $\D = \BB \YY$ are the delooping bicategories of monoidal categories $\X, \YY$ -- i.e., $\X$ and $\YY$ considered as bicategories with a single object -- then a lax functor $\C \to \D$ is the same thing as a lax monoidal functor $\X \to \YY$.

@@

\section{Lax natural transformations and functoriality of lax colimits}

@@

\section{Lax $2$-colimits in $\Cat$}

@@

\section{Examples of lax $2$-colimits in $\Cat$}

@@

This is significant, because it tells us that the category $\C/\X$ always satisfies a `factorization result' akin to those in \cite{SamsonGuyIAActive} (our Proposition \ref{PropInnocentFactorization}) and \cite{mcCHFiniteND}: if $f \from a \to b$ is a morphism in $\C/\X$, given by some morphism $\tilde{f}\from a \to x.b$ in $\C$, then we may write $f$ as the composite
\[
  a \xrightarrow{J(\tilde{f})} x.b \xrightarrow{\phi_{x,b}} b
  \]
in $\C$.  
This means that if we have a definability result for the category $\C$ -- for example, that every compact morphism in $\C$ is definable in some language $\L$ -- then we can automatically get a definability result for the category $\C/\X$ -- for example, that every compact morphism in $\C/\X$ (i.e., every morphism $a\to b$ in $\C/\X$ that is given by a compact morphism $a \to x.b$ in $\C$) is definable in the language $\L + \Phi_{x,a}$, where $\Phi_{x,a}$ is a new family of primitives in the language whose denotations are given by the $\phi_{x,a}$.
This is the same purpose that we used Proposition \ref{pKleisli} for when proving Theorem \ref{TheKleisliFullAbstraction}.

\section{Finite products distribute over lax colimits in $\Cat$}

@@

\section{Monoidal structure of $\C/\X$}

@@

\section{Symmetric monoidal structure of $\C/\X$}

@@

\section{Monoidal closed structure of $\C/\X$}

@@

\section{Cartesianness of the monoidal structure}

Suppose that a symmetric monoidal category $\X$ acts via a symmetric monoidal action on a Cartesian category $\C$.  
In this chapter we will consider what properties of the action we need in order to ensure that the induced monoidal structure on $\C/\X$ is Cartesian.

Since $\C$ is Cartesian, the diagonal and terminal maps in $\C$ gives every object a natural comonoid structure:
\begin{mathpar}
  a \xrightarrow{\Delta_a} a \times a
  \and
  a \xrightarrow{()} 1\,.
\end{mathpar}
Since the tensor product of objects in $\C/\X$ is given by the same object as in $\C$, we can lift this comonoid structure through the functor $J$ to give us a natural comonoid structure on each object of $\C/\X$:
\begin{mathpar}
  a \xrightarrow{J\Delta_a} a \tensor a
  \and
  a \xrightarrow{J()} I\,.
\end{mathpar}
We use the symbols $\tensor$ and $I$ rather than $\times$ and $1$ in $\C/\X$, since this tensor product is not necessarily Cartesian.

If every morphism $f\from a \to b$ in $\C/\X$ is a comonoid homomorphism with respect to this comonoid structure; i.e., if it makes the diagrams
\begin{mathpar}
  \begin{tikzcd}
    a \arrow[r, "f"] \arrow[d, "J\Delta_a"']
      & b \arrow[d, "J\Delta_b"] \\
    a \tensor a \arrow[r, "f\tensor f"]
      & b \tensor b
  \end{tikzcd}
  \and
  \begin{tikzcd}
    a \arrow[r, "f"] \arrow[d, "J()"']
      & b \arrow[dl, "J()"] \\
    I
  \end{tikzcd}
\end{mathpar}
commute, then we may identify $\C/\X$ with a full subcategory of its own category of comonoids, closed under tensor product.  
Since the category of comonoids in a monoidal category is automatically Cartesian, this will tell us that $\C/\X$ is Cartesian.

What does it mean for every morphism in $\C/\X$ to be a comonoid homomorphism?  
Firstly, since $\C$ itself is Cartesian, every morphism in $\C$ is a comonoid homomorphism with respect to the diagonal, and it follows that every morphism of the form $Jf \from a \to b$ (where $f\from a \to b$ is a morphism in $\C$) is a comonoid homomorphism in $\C/\X$.  
Since every morphism in $\C/\X$ may be written as the composite of a morphism of the form $Jf$ with a morphism of the form $\phi_{x,b}$, it will suffice to show that the morphisms of the form $\phi_{x,a}$ are comonoid homomorphisms in $\C/\X$; i.e., that they make the following diagrams commute.
\begin{mathpar}
  \begin{tikzcd}[column sep=40pt]
    x.a \arrow[r, "{\phi_{x,a}}"] \arrow[d, "J\Delta_{x.a}"']
      & a \arrow[d, "J\Delta_a"] \\
    x.a \tensor x.a \arrow[r, "{\phi_{x,a}\tensor\phi_{x,a}}"]
      & a \tensor a
  \end{tikzcd}
  \and
  \begin{tikzcd}
    x.a \arrow[r, "{\phi_{x,a}}"] \arrow[d, "J()"']
      & a \arrow[dl, "J()"] \\
    I
  \end{tikzcd}
\end{mathpar}
Let us compute the two arms of the first square directly in $\C$.  
The composite along the top right is given in $\C$ by the morphism
\[
  x.a \xrightarrow{x.\Delta_a}
  x.(a\times a)\,,
  \]
while that along the bottom right is given by the composite
\[
  x.a \xrightarrow{\Delta_{x.a}}
  x.a \times x.a \xrightarrow{m_{x,x,a,a}}
  (x\tensor x) . (a \times a)\,.
  \]
Then a sufficient condition for these two morphisms to be equal in $\C/\X$ is that for every object $x$ of $\X$ there should be a morphism $f \from x \to x\tensor x$ that makes the following diagram commute.
\[
  \begin{tikzcd}
    x.a \arrow[dr, "f.\Delta_a"] \arrow[d, "\Delta_{x.a}"']
      & \\
    x.a \times x.a \arrow[r, "{m_{x,x,a,a}}"]
      & (x\tensor x). (a\times a)
  \end{tikzcd}
  \]
Similarly, we can calculate the two arms of the second diagram directly in $\C$.  
The downwards arrow $J()$ is given by the composite
\[
  x.a \xrightarrow{()} 1 \xrightarrow{e_1} I.1\,,
  \]
while the composite $\phi_{x,a};J()$ is given by the composite
\[
  x.a \xrightarrow{x.()} x.1\,.
  \]
Thus, a sufficient condition for these two morphisms to be equal in $\C/\X$ is that for every object $x$ of $\X$ there should be a morphism $f_0 \from x \to I$ that makes the following diagram commute.
\[
  \begin{tikzcd}
    x.a \arrow[d, "()"] \arrow[dr, "f_0.()"']
      & \\
    1 \arrow[r, "e"]
      & I.1
  \end{tikzcd}
  \]

If $\X$ itself is Cartesian, then we have a natural choice of morphisms $x \to x \times x$ and $x \to 1$, but choosing $f$ to be the diagonal map and $g$ to be the terminal map does not automatically make the diagram above commute.  
For example, the Cartesian product is not the category-theoretic product in the category of sets and relations, even though this category is the Kleisli category for the symmetric monoidal monad $\powerset \from \Set \to \Set$, and may be identified with the category $\Set/1$, where $1$ and $\Set$ are both Cartesian.  
In this case, the diagram
\[
  \begin{tikzcd}
    \powerset(A) \arrow[dr, "\powerset(\Delta_A)"] \arrow[d, "\Delta_{\powerset(A)}"']
      & \\
    \powerset(A) \times \powerset(A) \arrow[r, "{m_{A,A}}" xshift=-3pt]
      & \powerset(A \times A)
  \end{tikzcd}
  \]
does not commute, the arrow at the top right sending a set $X\subseteq A$ to the set $\{(x,x) \suchthat x \in X\}$ while the composite at the bottom left sends $X$ to the set $\{(x,y)\suchthat x,y\in X\}$.

An alternative sufficient condition for $\C/\X$ to be Cartesian is for the morphisms mediating between the two arms of the squares to go in the other direction.  
In this case, we require that for each object $x$ of $\X$ there should be morphisms $g \from x\tensor x \to x$ and $g_0\from I \to x$ that make the following diagram commute.
\begin{mathpar}
  \begin{tikzcd}
    x.a \arrow[r, "x.\Delta_a"] \arrow[d, "\Delta_{x.a}"']
      & x.(a\times a) \\
    x.a \times x.a \arrow[r, "{m_{x,x,a,a}}"]
      & (x \tensor x).(a \times a) \arrow[u, "g.(a\times a)"']
  \end{tikzcd}
  \and
  \begin{tikzcd}
    x.a \arrow[r, "x.()"] \arrow[d, "()"']
      & x.1 \\
    1 \arrow[r, "e"]
      & I.1 \arrow[u, "g_0.1"']
  \end{tikzcd}
\end{mathpar}
This will be particularly useful for reader actions, where an oplax monoidal functor $\X\to \C$ gives rise to an action of $\oppcat \X$ on $\C$.  
Indeed, suppose that $j\from \X \to \C$ is an oplax symmetric monoidal functor between symmetric monoidal categories, where $\C$ is Cartesian closed.  
The functor $j$ induces a reader action of $\oppcat\X$ on $\C$ (i.e., $x.a = jx \to a$), and we want the category $\C/\oppcat\X$ to be Cartesian.  
For this, it will be sufficient to find, for each object $x$ of $\X$, morphisms $f \from x \to x\tensor x$ and $f_0 \from x \to I$ such that the following diagrams commute.
\begin{mathpar}
  \begin{tikzcd}
    jx \to a \arrow[rr, "jx \to \Delta_a"] \arrow[d, "\Delta_{jx \to a}"']
      &[-10pt]
        &[+10pt] jx \to (a \times a) \\
    (jx \to a) \times (jx \to a) \arrow[r]
      & (jx \times jx) \to (a \times a) \arrow[r, "m^j_x \to (a\times a)"]
        & j(x\tensor x) \to (a \times a) \arrow[u, "jf \to (a \times a)"']
  \end{tikzcd}
  \and
  \begin{tikzcd}
    jx \to a \arrow[rr, "jx \to ()"] \arrow[d, "()"']
      &
        & jx \to 1 \\
    1 \arrow[r]
      & 1 \to 1 \arrow[r, "e^j \to 1"]
        & jI \to 1 \arrow[u, "jf_0 \to 1"']
  \end{tikzcd}
\end{mathpar}
For these to commute, it is sufficient that the composite
\[
  jx \xrightarrow{jf} j(x \tensor x) \xrightarrow{m^j_{x,x}} jx \times jx
  \]
should be equal to the diagonal on $jx$, and that the composite
\[
  jx \xrightarrow{jf_0} jI \xrightarrow{e^j} 1
  \]
should be equal to the terminal morphism $jx \to 1$.  
Of course, this second condition is automatically satisfied by the definition of a terminal object, so all we require is that some morphism $f_0 \from x \to I$ should exist.

We have proved:

@@

  Suppose that for every object $x$ of $\X$, there are morphisms $f \from x \to x\tensor x$ and $f_0 \from x \to I$ in $\X$ such that $jf;m^j_{x,x}=\Delta_{jx}$.
  Then $\C/\oppcat \X$ is Cartesian closed.
  \label{TheCartesianClosedCx}
\end{theorem}

In the next chapter, we will investigate a class of reader actions, and give a more concrete property for the category $\C/\oppcat\X$ to be Cartesian closed.

\chapter{Reader actions on $\Set$}

\section{Colimits of actions are monoidal functors}

@@

        \colim_{t,z\from \X}\limits t.(a \tensor b) \tensor z.c \arrow[ddr, "{\Colim m_{t,z,a\tensor b,c}}" description, to=L.west, bend right=30]

@@

              \& \colim_{x,t\from \X}\limits x.a \tensor t.(b\tensor c) \arrow[ddl, "{\Colim m_{x,t,a,b\tensor c}}" description, to=R.east, bend left=30] \\
          \& \colim_{x,y,z\from\X}\limits ((x\tensor y) \tensor z).((a \tensor b) \tensor c) \arrow[r, "\Colim\assoc.\assoc" yshift=5pt] \arrow[d, hookrightarrow]
            \& \colim_{x,y,z\from\X}\limits (x \tensor (y\tensor z)).(a \tensor (b \tensor c)) \arrow[d, hookrightarrow]
          \& |[alias=L]| \colim_{t,z\from \X}\limits (t\tensor z).((a \tensor b)\tensor c) \arrow[d, hookrightarrow]
            \& |[alias=R]| \colim_{x,t\from \X}\limits (x \tensor t).(a \tensor (b \tensor c)) \arrow[d, hookrightarrow]
          \& \colim_{u\from \X}\limits u.((a\tensor b) \tensor c) \arrow[r, "\Colim u.\assoc"]

@@

\section{Reader actions on $\Set$ vs change of base}

@@

\begin{proposition}
  Let $j \from \X \to \Set$ be an oplax symmetric monoidal functor between symmetric monoidal categories.
  Let $\oppcat\X$ act on $\Set$ via the reader action induced by $j$.
  Then the \Mellies category $\Mell_{\oppcat\X}\Set$ is isomorphic to the category obtained via base change along the action regarded as a functor
  \[
    \blank.\blank\from \Set \to [\X,\Set]\,.
    \]
  \label{PropMelliesVsBaseChange}
\end{proposition}
\begin{proof}
  The objects of both categories are sets.  
  The morphism objects in the \Mellies category are given by
  \[
    \Mell_\X\Set(A,B)(x) = [A,[jx,B]]\,,
    \]
  while those in the base-changed category are given by
  \[
    (\blank.\blank)_*\Set(A,B)(x) = [jx, [A,B]]\,,
    \]
  and these may be related by the symmetry isomorphism
  \[
    s_{jx,A,B} \from [A,[jx,B]] \to [jx, [A,B]]\,.
    \]
  We need to show that this preserves composition of morphisms.  
  Recall that composition in the \Mellies category is given by
  \begin{IEEEeqnarray*}{Cl}
    & \int^{y,z} [A,[jy,B]] \times [B,[jz,C]] \times \X(y\tensor z,x) \\
    \to &
    \int^{y,z} [A,[j(y\tensor z),C]]\times \X(y\tensor z,x) \\
    \cong &
    [A,[jx,C]]\,,
  \end{IEEEeqnarray*}
  where the first arrow is induced via the \Mellies composition, while composition in the base changed category is given by
  \begin{IEEEeqnarray*}{Cl}
    &\int^{y,z}[jy,[A,B]] \times [jz,[B,C]] \times \X(y\tensor z,x) \\
    \xrightarrow{\makebox[144pt]{$\int^{y,z} m_{jy,jz,[A,B],[B,C]}\times \X(y\tensor z,x)$}} &
    \int^{y,z}[jy \times jz,[A,B]\times [B,C]] \times \X(y,\tensor z,x) \\
    \xrightarrow{\makebox[144pt]{$\int^{y,z} [m^j_{y,z}, ;] \times \X(y\tensor z,x)$}} &
    \int^{y,z}[j(y \tensor z),[A,C]]\times \X(y\tensor z,x) \\
    \cong & 
    [jx,[A,C]]\,.
  \end{IEEEeqnarray*}
  where $;$ is the internal composition in $\Set$.

@@

\section{From monoidal endofunctors to reader actions}

@@

Now recall that in order to apply the second part of Theorem \ref{TheCartesianClosedCx}, and deduce that $\Set/\oppcat\X$ is Cartesian closed, we must prove that for every object $x$ of $\X$, there are morphisms $f \from x \to x \tensor x$ and $f_0 \from x \to I$ such that $jf = \Delta_{jx};m^j_x$ and $jf_0 = ();e^j$.
This is particularly useful in our case, since the functor $\X \to \Set$ is faithful, making $\X$ into a concrete category.  
Then the morphisms between two objects in $\X$ may be thought of as `structure-preserving functions' between the corresponding sets.  
The criterion from Theorem \ref{TheCartesianClosedCx} then says that the composites $\Delta_{jx};m^j_x$ and $();e^j$ are structure-preserving.

In other words, for every set $A$ and every $p\in FA$, the function $\Delta_A$ must define a morphism
\[
  (A,p) \to (A \times A, m^F_{A,A}(p,p))\,\text{--}
  \]
i.e., we must have
\[
  F(\Delta_A)(p) = m^F_{A,A}(p,p)\,\text{--}
  \]
and the function $()$ must define a morphism
\[
  (A,p) \to (1, e^F)\,\text{--}
  \]
i.e., we must have
\[
  F(())(p) = e^F\,.
  \]
Let us see what this means in some examples.
\begin{example}
  The finite powerset functor does not satisfy the condition given above; indeed, we have
  \begin{mathpar}
    \powerset_+(\Delta_A)(X) = \{(x,x)\in A \times A \suchthat x \in X\}
    \and
    m^{\powerset_+}_{A,A}(X,X) = X \times X = \{(x,y) \in A \times A \suchthat x \in X\}\,.
  \end{mathpar}
  Similarly, the discrete probability measure functor does not satisfy the condition we have given, since the diagonal map
  \[
    \Delta_A \from (A,\bP_A) \to (A \times A, \bP_{A\times A})
    \]
  does not preserve probability in general.
\end{example}
